A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-29 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: two typos
Two more obvious typos I missed.

TODO: figure with classical triangle and quantum Bloch Ball

@@

The state capacity, then, recovers the number of distinguishable states. In classical mechanics, all states are perfectly distinguishable as they represent mutually exclusive cases, the measure is additive. In quantum mechanics, only orthogonal states are distinguishable, therefore the measure is non-additive over pure states. We see again how the lack of single decomposition of ensembles over pure states, which equates to the lack of mutually exclusiveness, is responsible for the failure of the additivity of the measure. Note that, since the fraction capacity depends only on the ensemble space, and not also on a chosen ensemble, the fraction capacity essentially characterizes the shape of the ensemble space itself.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\columnwidth]{ChoquetCalculation2.jpg}
	\caption{...}
\end{figure}

@@

To recap, we have (TODO: add a graph)
\begin{equation}
	\frcap_{\ens[o]}(A(s))=
	\begin{cases}
		1 & \text{if } s\leq \frac{1}{2} a + \frac{1}{2} b\\
		\frac{1}{2} \frac{b - a}{s-a} & \text{if } \frac{1}{2} a + \frac{1}{2} b < s \leq b\\
		0 & \text{if } s > b.
	\end{cases}
\end{equation}
We can now calculate the integral
\begin{equation}
	\begin{aligned}
		(C)&\int F d\frcap_{\ens[o]} := 
		\int_{-\infty}^0
		(\frcap_{\ens[o]} (A(s))-\frcap_{\ens[o]}(X))\, ds \\
		&+
		\int^\infty_0
		\frcap_{\ens[o]} (A(s))\, ds \\
		&= \int_{-\infty}^0 (1 - 1)ds + \int_{0}^{\frac{1}{2} a + \frac{1}{2} b}  1 ds \\
		&+ \int_{\frac{1}{2} a + \frac{1}{2} b}^{b}  \frac{1}{2} \frac{b - a}{s-a} ds + \int_{b}^{\infty} 0 ds \\
		&=\left[s\right]_{0}^{\frac{1}{2} a + \frac{1}{2} b}
		+ \frac{1}{2}\left(b - a\right)\left[\ln (s - a)\right]_{\frac{1}{2} a + \frac{1}{2} b}^{b} \\
		&= \frac{1}{2} a + \frac{1}{2} b + \frac{1}{2}\left(b - a\right)\left[\ln (b - a)\right. \\
		&+\left. \ln \left(\frac{1}{2} a + \frac{1}{2} b - a\right) \right] \\
		&= \frac{1}{2} a + \frac{1}{2} b + \frac{1}{2}\left(b - a\right)\ln (2)
	\end{aligned}
\end{equation}

>>>>

	% dimensions
	\newcommand{\radius}{1}

	\begin{tikzpicture}[x={(\xx cm, \xy cm)}, y={(\yx cm, \yy cm)}, z={(\zx cm, \zy cm)}, scale=1.2]
		% Triangle
		\draw (canvas cs: x=-3cm, y=-0.25cm) +(canvas polar cs: angle=90, radius=1cm) -- +(canvas polar cs: angle=210, radius=1cm) -- +(canvas polar cs: angle=330, radius=1cm) -- cycle;

		% Bloch sphere (outline drawn in canvas plane)
		\draw (0, 0, 0) circle (\radius cm);

		% meridians and equator
		\begin{scope}[canvas is yz plane at x=0]
			\draw (\ixy, \ixz) arc[start angle={atan2(\ixz, \ixy)}, delta angle=180, radius=1 cm];
			\draw[dashed] (-\ixy, -\ixz) arc[start angle={atan2(-\ixz, -\ixy)}, delta angle=180, radius=1 cm];
		\end{scope}
		\begin{scope}[canvas is zx plane at y=0]
			\draw[dashed] (\iyz, \iyx) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=1 cm];
			\draw (-\iyz, -\iyx) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=1 cm];
		\end{scope}
		\begin{scope}[canvas is xy plane at z=0]
			\draw[dashed] (\izx, \izy) arc[start angle={atan2(\izy, \izx)}, delta angle=180, radius=1 cm];
			\draw (-\izx, -\izy) arc[start angle={atan2(-\izy, -\izx)}, delta angle=180, radius=1 cm];
		\end{scope}
	\end{tikzpicture}
	\caption{Examples of a classical discrete ensemble space with three pure states (triangle) and of a two dimensional quantum ensemble space (ball).}
\end{figure}

@@

The state capacity, then, recovers the number of distinguishable states. In classical mechanics, all states are perfectly distinguishable as they represent mutually exclusive cases, the measure is additive. In quantum mechanics, only orthogonal states are distinguishable, therefore the measure is non-additive over pure states. We see again how the lack of single decomposition of ensembles over pure states, which equates to the lack of mutual exclusiveness, is responsible for the failure of the additivity of the measure. Note that, since the fraction capacity depends only on the ensemble space, and not also on a chosen ensemble, the fraction capacity essentially characterizes the shape of the ensemble space itself.

@@

	% dimensions
	\newcommand{\s}{0.5}
	\pgfmathsetmacro{\radiusats}{sqrt(1 - \s*\s)}
	\pgfmathsetmacro{\stheta}{asin(\s)}
	\pgfmathsetmacro{\minstheta}{asin(-\s)}

	\begin{tikzpicture}[x={(\xx cm, \xy cm)}, y={(\yx cm, \yy cm)}, z={(\zx cm, \zy cm)}, scale=1.2]
		% Bloch sphere (outline drawn in canvas plane)
		\draw (\xy, \yy, \zy) arc[start angle=90, end angle={acos(\radiusats*\iyx*\xx - \s*\yx + \radiusats*\iyz*\zx)}, radius=1 cm] coordinate (A);
%		\draw[dash dot, red] (A) arc[start angle={acos(\radiusats*\iyx*\xx - \s*\yx + \radiusats*\iyz*\zx)}, end angle=-90, radius=1 cm];
		\draw (\xy, \yy, \zy) arc[start angle=90, end angle={acos(-\radiusats*\iyx*\xx - \s*\yx - \radiusats*\iyz*\zx)}, radius=1 cm] coordinate (B);
%		\draw[dash dot, green] (B) arc[start angle={acos(-\radiusats*\iyx*\xx - \s*\yx - \radiusats*\iyz*\zx)}, end angle=270, radius=1 cm];
		\draw[dashed] (A) arc[start angle={acos(\radiusats*\iyx*\xx - \s*\yx + \radiusats*\iyz*\zx)}, end angle={acos(-\radiusats*\iyx*\xx - \s*\yx - \radiusats*\iyz*\zx)-360}, radius=1 cm];

		% special points
		\shadedraw (0, 0, 0) circle (0.03 cm) node[below right] {$\mathsf{o}$};
		\shadedraw (0, \radius, 0) circle (0.03 cm) node[above] {$\mathsf{b}$};
		\shadedraw (0, -\radius, 0) circle (0.03cm) node[below] {$\mathsf{a}$};
		\shadedraw (0, \s, 0) circle (0.03cm) node[left] {$\mathsf{e}$};

		% convex combination
		\draw (0, -\radius, 0) -- (0, \s, 0);
		\draw (2.25*\xx, 2.25*\yx, 2.25*\zx) node {$\ens[o] = \frac{\overline{\ens[o]\ens[a]}}{\overline{\ens[e]\ens[a]}} \ens[e] + \frac{\overline{\ens[o]\ens[e]}}{\overline{\ens[e]\ens[a]}} \ens[a]$};

		% s axis
		\begin{scope}[shift={(-1.5*\xx, -1.5*\yx, -1.5*\zx)}]
			\draw (0, -1, 0) -- (0, 0, 0);
			\draw[->] (0, 0, 0) -- (0, 1, 0) node[below left] {$s$};
		\end{scope}

		% cut
		\begin{scope}[canvas is zx plane at y=\s]
%			\filldraw[color=blue, thick, draw opacity=1, fill opacity=0.1] (-\margin, -\margin) -- (\margin, -\margin) -- (\margin, \margin) -- (-\margin, \margin) -- cycle;
			\draw[blue, thick, dashed] ({\iyz * \radiusats}, {\iyx * \radiusats}) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=\radiusats cm];
			\draw[blue, thick] ({-\iyz * \radiusats}, {-\iyx * \radiusats}) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=\radiusats cm];
		\end{scope}

		% meridians and equator
		\begin{scope}[canvas is yz plane at x=0]
			% foreground above s
			\draw (\ixy, \ixz) arc[start angle={atan2(\ixz, \ixy)}, end angle=90-\stheta, radius=1 cm] coordinate (C);
			% foreground below s
			\draw[dashed] (C) arc[start angle=90-\stheta, end angle={atan2(-\ixz, -\ixy)}, radius=1 cm] coordinate (D);
			% background below s
			\draw[loosely dotted] (D) arc[start angle={atan2(-\ixz, -\ixy)}, end angle=\stheta+270, radius=1 cm];
			% background above s
			\draw[densely dotted] (\ixy, \ixz) arc[start angle={atan2(\ixz, \ixy)}, end angle=\stheta-90, radius=1 cm];
		\end{scope}
%		\begin{scope}[canvas is xy plane at z=0]
%			\draw[dashed] (\izx, \izy) arc[start angle={atan2(\izy, \izx)}, delta angle=180, radius=1 cm];
%			\draw (-\izx, -\izy) arc[start angle={atan2(-\izy, -\izx)}, delta angle=180, radius=1 cm];
%		\end{scope}
		\begin{scope}[canvas is xy plane at z=0]
			% foreground above s
			\draw (-\izx, -\izy) arc[start angle={atan2(-\izy, -\izx)}, end angle=180-\stheta, radius=1 cm] coordinate (E);
			% foreground below s
			\draw[dashed] (E) arc[start angle=180-\stheta, end angle={atan2(\izy, \izx)+360}, radius=1 cm] coordinate (F);
			% background below s
			\draw[loosely dotted] (F) arc[start angle={atan2(\izy, \izx)}, end angle=\stheta, radius=1 cm];
			% background above s
			\draw[densely dotted] (-\izx, -\izy) arc[start angle={atan2(-\izy, -\izx)}, end angle=\stheta, radius=1 cm];
		\end{scope}
	\end{tikzpicture}
	\caption{Geometric representation of the element $\ens[e]$ in $A(s)$ with the largest fraction of $\ens[o]$.}
\end{figure}

@@

To recap, we have
\begin{equation}
	\frcap_{\ens[o]}(A(s))=
	\begin{cases}
		1 & \text{if } s\leq \frac{1}{2} a + \frac{1}{2} b\\
		\frac{1}{2} \frac{b - a}{s-a} & \text{if } \frac{1}{2} a + \frac{1}{2} b < s \leq b\\
		0 & \text{if } s > b.
	\end{cases}
\end{equation}
\begin{figure}[H]
	\centering
	\newcommand{\ticklength}{0.03}
	\begin{tikzpicture}[scale=3, xticklabel/.style={anchor=base, inner sep=1pt, yshift=-2.5ex}]
		% axes
		\draw[->] (0, 0) -- (1.7, 0);
		\draw (0, 0) node[left] {$0$} -- +(-\ticklength, 0) (0, 0.25) node[left] {$1/2$} -- +(-\ticklength, 0) (0, 0.5) node[left] {$1$} -- +(-\ticklength, 0);
		\draw[->] (0, 0) -- (0, 0.6);
		\draw (0, 0) node[xticklabel] {$0$} -- +(0, -\ticklength) (0.5, 0) node[xticklabel] {$a$} -- +(0, -\ticklength) (1, 0) node[xticklabel] {$1/2(a+b)$} -- +(0, -\ticklength) (1.5, 0) node[xticklabel] {$b$} -- +(0, -\ticklength);
		% cases s <= 1/2(a+b) and 1/2(a+b) < s <= b
		\draw[thick, blue, domain=1:1.5, samples=200] (0, 0.5) --plot (\x, {0.25/(\x-0.5)});
		% mark discontinuity
		\filldraw (1.5, 0.25) circle (0.02cm);
		\draw (1.5, 0) circle (0.02cm);
		% case s > b
		\draw[thick, blue] (1.5, 0) -- (1.7, 0);
	\end{tikzpicture}
	\caption{The integrand $\frcap_{\ens[o]}(A(s))$ as a function of $s$.}
\end{figure}
We can now calculate the integral
\begin{equation}
	\begin{aligned}
		(C)&\int F d\frcap_{\ens[o]} := 
		\int_{-\infty}^0
		(\frcap_{\ens[o]} (A(s))-\frcap_{\ens[o]}(X))\, ds \\
		&+
		\int^\infty_0
		\frcap_{\ens[o]} (A(s))\, ds \\
		&= \int_{-\infty}^0 (1 - 1)ds + \int_{0}^{\frac{1}{2} a + \frac{1}{2} b}  1 ds \\
		&+ \int_{\frac{1}{2} a + \frac{1}{2} b}^{b}  \frac{1}{2} \frac{b - a}{s-a} ds + \int_{b}^{\infty} 0 ds \\
		&=\left[s\right]_{0}^{\frac{1}{2} a + \frac{1}{2} b}
		+ \frac{1}{2}\left(b - a\right)\left[\ln (s - a)\right]_{\frac{1}{2} a + \frac{1}{2} b}^{b} \\
		&= \frac{1}{2} a + \frac{1}{2} b + \frac{1}{2}\left(b - a\right)\left[\ln (b - a)\right. \\
		&+\left. \ln \left(\frac{1}{2} a + \frac{1}{2} b - a\right) \right] \\
		&= \frac{1}{2} a + \frac{1}{2} b + \frac{1}{2}\left(b - a\right)\ln (2)
	\end{aligned}
\end{equation}
